Personalised Map Matched Imputation: imputing missing data from smartphone location logs

### 2018-05-08 Thesis Manuscript/manuscript.Rmd
final commit

  Submitted as a research report conforming to APA manuscript guidelines (6th edition).

abstract: |
  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health. The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data. Nonetheless, smartphone collected personal mobility data has remained underused in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it. Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain. The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.  This results in a high proportion of missing data and significant noise. In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs. The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data. In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurements and impute missing data values. By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals. When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.

@@

class             : "man"
output            : papaja::apa6_pdf

@@

Why is human mobility important? Human mobility measures are quantified metrics of how people move about in their environment. Human mobility affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@goldberg_international_2006]. At their root these questions are mobility measures. In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g. @villanueva_association_2013]. Economists are interested not only in where people live, but also where they go to work: geographic labour mobility is related to the income of individuals, the well-being of a country's economy and even informs the policy of bodies such as the European Comission [@tatsiramos_geographic_2009]. The extensive use of measures like these across different domains in social science strongly suggests that mobility metrics are linked to important real world outcomes. Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017]. Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough. When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary across environments.

Despite the importance of mobility measures, the majority of these metrics are obtained through the use of questionnaires (such as the aforementioned personality questionnaire) or specifically through pen-and-paper travel diaries. As a thought experiment we encourage the reader to try to remember where exactly he or she was five weeks ago on Saturday at midday. Most people find this sort of task very difficult (if not impossible), which illustrates that questionnaires and travel diaries are have well-known methodological flaws. This method of collecting data is difficult and relies on accurate self-reporting. Travel diaries are burdensome to collect because participants must be explicitly asked to write down on their movement patterns at frequent intervals due to human forgetfulness. This makes the data collection expensive and limits the duration of data collection in practice. In addition, the frequent reporting duties of the participant may bias the participants behaviour. The limits of human cognition also limit the accuracy of self-reported measures. There is evidence that participants are systematically biased when self-reporting mobility measures. For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017]. These obstacles can only be overcome by using objective data on human mobility.

Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones. Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times. @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour. Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016]. Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers. Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times. These smartphones can collect and store hundreds of location measurements a day. For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017]. Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017]. Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millions of individuals spanning several years at little cost and without a significant burden of the participants.

@@

Researchers often develop custom-made tracking applications which participants are instructed to download on their phone. Alternatively participants are given a phone to use for a given period of time with the custom-made tracking app pre-installed. We call location logs resulting from these custom-made apps *custom logs*. The advantage of custom logs is that the researchers can adjust tracking parameters, such as the the frequency of measurements and the sensor with which they are made. The disadvantage with this approach is that researchers have to develop or adapt a custom-made tracking application (which is not easy given hundreds of different types of smartphone models), distribute it among research participants and enforce participation. Participants may dislike tracking apps because they view them as more intrusive and these apps regularly drain the battery of the device [@harari_using_2016]. Moreover, researchers have distribute this application among research participants and convince them not to turn it off.

We focus on another solution, which is to take advantage of existing smartphone location logs  (*secondary logs*) . The advantages are clear: repositories such as Google's Location History contains information on millions of users spanning years [@location_history_timeline_2017], participants can share the data by the click of a button and there can be no behavioural changes due to participation in the study as the participant share past data. The disadvantage is that researchers have no control over the tracking parameters, often resulting in logs with sparse and inaccurate measurements. Hence, two important challenges are dealing with missing data and measurement noise.

In order to work with secondary logs, researchers need to be able to handle the data sparsity that leads to missing data. Missing data is a pervasive issue in secondary logs as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home or switching the device off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in different fields [e.g. @harari_using_2016;@jankowska_framework_2015]. In Figure \@ref(fig:longMeasurementsPerDay) shows that despite the long duration of the log the sparsity it is also evident.

@@

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are auto-correlated in both time and space. This means that best practices with other types of data, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work, even though she has never been there. Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS location trackers because smartphones often use less accurate sensors. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time. Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location. Most notably, low accuracy is often (but not always) associated with movement (see Figure \@ref(fig:accuracyPlot2)).

@@

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure \@ref(fig:palmiusVsMe1). The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs. We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simulated as missing and their imputed  counterparts. Finally we will calculate time spent at home as a function of the imputation method.

@@

How have researchers dealt with missing data in human mobility logs thus far? Unfortunately there is no golden standard in how to deal with this type of missing data. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons_false-positive_2011]. The vagueness is possibly also due to the fact that most researchers are unfamiliar with possible solutions. Most researchers simply down-sample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g. @palmius_detecting_2017). The only principled approach that we know of that aims to solve the issue of missing data in location logs as they relate to human mobility is that of @barnett_inferring_2016. We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing exploring other spatiotemporal methods.

A lack in methods for missing data imputation for human mobility patterns does not imply there is not a vast literature on modelling movement. The most widespread models are SSMs, therefore we shall detail a few examples and subsequently argue that they are nonetheless unsuited for long term human mobility logs. Ecologists have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008]. These models can be quite complex. @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013]. The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.

For secondary logs the main limitation of SSM implementations is that they ignore movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. In other words, the estimated location $G(t)$ at time-point $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. Moreover, these models often assume time and space invariance (location is not a direct function of time or space). These mathematical assumptions are violated in the case of human movement patterns. For this reason we do not consider existing SSMs to be useful for imputing missing data in this case.

@@

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @@wu_spatial-temporal-semantic_2017]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measurement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

>"If the standard deviation of [$D$] in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch".

Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ after 9pm. In cases where the previous conditions are not met no values are imputed.

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only principled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. In the words of the authors, @barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses. Flights are segments of linear movements and pauses corresponding to periods of time where a person does not move. Subsequently, the authors impute missing data by:

@@

Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidean plane. Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. For simplicity, we can assume that the x-axis is the longitude and the y-axis is the latitude. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the latitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data. Furthermore, $G(t)$, $D$ and $a$ are indexed by time labeled by the countable set $t = t_1 < ... < t_{n+1}$. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

When several data sets are available from individuals living in overlapping areas we can construct a $t \times i$ matrix $M$ where the entry $M(t,i)$ contains $G(t)$ for the individual $i$.

## Personalised Map Matching Imputation

@@

First, following @barnett_inferring_2016 we categorise all time-points $t$ as either belonging to the set $P$ (pause) or set $F$ (flight). Conceptually pauses can be understood as periods of time where an individual spends significant amount of continuous time without moving. Flights are the times where the individual is moving. Let $t_a$ be a pause of length $n$.
$$t_a = t_i < ... < t_{i+n}$$
Let $t_b$ be a pause of length $m$ such that there is no temporal overlap between $t_a$ and $t_b$:
$$t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j$$
Then it follows that between the two pauses there must be a flight indexed by $t_x$ of length $j-i+n$.
$$t_x =  t_{i+n} < ... < t_{j} |t_x \in F$$
We define pause locations $G(t_a), G(t_b) | t_a, t_b \in P$ as locations where an individual spends an extended amount of time in the same space (e.g. school, home, work, train station, barber shop, bar, gym). Importantly, our model assumes period and cyclic human movement such that there are many pauses $t_{a1},t_{a2},...,t_{an}$ such that $G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})$. Moreover,it is possible for $G(t_a) = G(t_b)$ such that $t_a \ne t_b$. For example, if the individual leaves home for a run and returns home without stopping anywhere else.

@@

## Personalised Map Matching Imputation algorithm

@@

1. *Map building*: Extract from measurements $D$ all pause location bins and path location bins to create a personalised map.
2. *Binning*: Assign each measurement $D$ to a unique discrete location bin. 
3. *Imputing*: Use a classification method to predict missing measurements based on all the available information.

@@

Following @wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

@@

1. The next measurement $D_{t+1}$ is within $t_{\text{Pause lim}}$ amount of seconds (so it is not missing)
2. The next measurement $D_{t+1}$ is within $d_{\text{Pause lim}}$ meters.
3. The duration of the pause is more than $\delta_{\text{Pause lim}}$ seconds.
4. Let the measurements of a possible pause which fit the aforementioned criteria be $D_{t,t+1,...,t+n}$. These points are only a pause if the distance between the mean coordinates of $D_{t,t+1,...,t+n}$ and the furthest away points of $D_{t,t+1,...,t+n}$ is within 2 times the mean accuracy $a$ of $D_{t,t+1,...,t+n}$.

This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a bin.

@@

1. Take all measurements which are not pauses, filter them based on an accuracy threshold $a_{\text{Path Lim}}$.
2. Create a distance matrix for all remaining measurements $D_t \in F$ and hierarchically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.

At this point all empirically observed path bins and pause bins are extracted. However, there may be some overlap between pause bins and path bins. Thus, the bins are clustered again, such that the pause bins retain priority. This means that if a pause bin and a path bin are within less than $d$ meters of each other, the path bin is removed. The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster. The end result is a discretised map which contains pause and flight bins based on the entire log history of the individual.

@@

PMMI uses a similar logic, but without using any external geographic database. The key modification in PMMI is that whilst @wu_spatial-temporal-semantic_2017 uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map. This map is subsequently used to bin measurements. This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long. To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_{ab}$.

Thus, all measurements $D$ were are then assigned to a discrete bin on the personal map. This includes previous measurements which were discarded from the map building exercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$. In this implementation we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.

>>>>

  Submitted as a thesis manuscript conforming to APA manuscript guidelines (6th edition).

abstract: |
  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and health. The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data. Nonetheless, smartphone collected personal mobility data has remained underused in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it. Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain. The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.  This results in a high proportion of missing data. In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs. The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data. In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurements and impute missing data values. By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals. When imputing a subset of real missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.

@@

class             : "man"
output            : papaja::apa6_pdf
keep_tex: true

@@

Why is human mobility important? Human mobility affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. Human mobility measures are quantified metrics of how people move about in their environment. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@goldberg_international_2006]. At their root these questions are mobility measures. In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g. @villanueva_association_2013]. Economists are interested not only in where people live, but also where they go to work: geographic labour mobility is related to the income of individuals, the well-being of a country's economy and even informs the policy of bodies such as the European Comission [@tatsiramos_geographic_2009]. The extensive use of measures like these across different domains in social science strongly suggests that mobility metrics are linked to important real world outcomes. Behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017]. To fully understand behaviour not only must we understand how behaviour can vary within and between individuals, but also across environments.

Despite the importance of mobility measures, the majority of  metrics are obtained through the use of questionnaires (such as the aforementioned personality questionnaire) or specifically through pen-and-paper travel diaries. As a thought experiment we encourage the reader to try to remember where exactly he or she was five weeks ago on Saturday at midday. Most people find this sort of task difficult (if not impossible), which illustrates the main methodological flaws of questionnaires and travel diaries. Travel diaries are burdensome to collect because participants must be explicitly asked to write down on their movement patterns at frequent intervals (lest they forget them). The burden to the participants makes data collection expensive, which in turn limits the duration of data collection in practice. In addition, the frequent reporting duties of the participant may bias the participant's behaviour. The limits of human cognition also limit the accuracy of self-reported measures. There is evidence that statistics derived from self-reported mobility measures are systematically biased. For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017]. These obstacles can only be overcome by using objective data on human mobility.

Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones. Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times. @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour. Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016]. Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers. Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times. These smartphones can collect and store hundreds of location measurements a day. For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017]. Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017]. Researchers now have the ability to easily access the objective movement records of millions of individuals spanning several years at little cost and without a significant burden to the participants.

@@

Researchers often develop custom-made tracking applications which participants are instructed to download on their phone. Alternatively participants are given a phone to use for a given period of time with the custom-made tracking app pre-installed. We call location logs resulting from these custom-made apps *custom logs*. The advantage of custom logs is that the researchers can adjust tracking parameters, such as the the frequency of measurements and the sensor used to make the measurement. The disadvantage of this approach is that researchers have to develop or adapt a custom-made tracking application (which is not easy given hundreds of different types of smartphone models), distribute it among research participants and enforce participation. Participants may dislike tracking apps because they view them as more intrusive and these apps regularly drain the battery of the device [@harari_using_2016].

We focus on another solution, which is to take advantage of existing smartphone location logs  (*secondary logs*). The advantages are clear: repositories such as Google's Location History contain years of information on millions of users [@location_history_timeline_2017] and participants can share the data by clicking of a button. The disadvantage is that researchers have no control over the tracking parameters, often resulting in logs with sparse and inaccurate measurements. Hence, two important challenges in using secondary logs are how to deal with missing data and measurement noise.

In order to work with secondary logs, researchers need to be able to handle the data sparsity that leads to missing data. Missing data is a pervasive issue in secondary logs as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include switching the device or its location services off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in different fields [e.g. @harari_using_2016;@jankowska_framework_2015]. Figure \@ref(fig:longMeasurementsPerDay) shows a secondary log spanning years where despite its long duration the sparsity of measurements is also evident.

@@

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are auto-correlated in both time and space. This means that common practices with other types of data, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work, even though she has never been there. Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Another methodological problem is related to the noise in the measurements. The accuracy of smartphone location measurements is substantially lower than that of professional GPS location trackers because smartphones often use less accurate sensors. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time. Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location. Most notably, low accuracy is often (but not always) associated with movement (see Figure \@ref(fig:accuracyPlot2)).

@@

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distances between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs. PPMI creates a discrete personalised map based on the individuals mobility history. Then it assigns each measurement to a discrete location based on the individuals entire mobility history. Finally, it uses a classification method to impute missing periods.

@@

How have researchers dealt with missing data in human mobility logs thus far? Unfortunately there is no golden standard in how to deal with this type of missing data. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons_false-positive_2011]. Most researchers simply down-sample temporally and remove missing observations or use some sort of rule-based common sense imputations [e.g. @palmius_detecting_2017]. The only principled approach (that we know of) aiming to solve the issue of missing data in this context (i.e. location logs as they relate to human mobility) is that of @barnett_inferring_2016. We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing other spatiotemporal methods.

A lack in methods for missing data imputation for human mobility patterns does not imply there is not a vast literature on modelling movement. The most widespread models are SSMs, therefore we shall detail a few examples and subsequently argue that they are nonetheless unsuited for long term human mobility logs. Ecologists have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008]. @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013]. The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.

For secondary logs the main limitation of SSM implementations is that they ignore movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. In other words, the estimated location at a time-point $t$ is often based only upon measurements at timepoints $t$ and ${t-1}$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. Moreover, these models often assume time and space invariance (i.e. location is not a direct function of time or space). These mathematical assumptions are violated in the case of human movement patterns. For this reason we do not consider existing SSMs to be useful for imputing missing data in smartphone mobility logs.

@@

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @wu_spatial-temporal-semantic_2017]. Importantly, none of them worked with secondary logs, but rather with custom logs that have higher sampling frequency. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measurement inaccuracy in custom logs by removing from the data set all unique low-accuracy data points with a speed greater than 100 km/h. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

>"If the standard deviation of in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch".

Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ if this occurred after 9pm. In cases where the previous conditions are not met no values are imputed.

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only principled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. @barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses. Flights are segments of linear movements and pauses corresponding to periods of time where a person does not move. Subsequently, the authors impute missing data by:

@@

Subsequently let us simplify by assuming that a persons location is on a two-dimensional Euclidean plane. Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. For simplicity, we can assume that the x-axis is the longitude and the y-axis is the latitude. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the latitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data. Furthermore, $G(t)$, $D$ and $a$ are indexed by time labeled by the countable set $t = t_1 < ... < t_{n+1}$. If $D_t = \emptyset$ it is considered *missing*. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. The reader should keep in mind that a high accuracy value $a_t$ for a measurement $D_t$ means that the measurement is inaccurate (this can be rather confusing).

## Personalised Map Matched Imputation

@@

First, following @barnett_inferring_2016 we categorise all time-points $t$ as either belonging to the set $P$ (pause) or set $F$ (flight). Conceptually pauses can be understood as periods of time where an individual spends significant amount of continuous time without moving. Flights are the times where the individual is moving. Let $t_a$ be a pause of length $n$.
$$t_a = t_i < ... < t_{i+n}$$
Let $t_b$ be a pause of length $m$ such that there is no temporal overlap between $t_a$ and $t_b$ and $t_b$ is the first pause after the end of $t_a$:
$$t_b = t_j < ... < t_{j+m}|t_{i+n}< t_j$$
Then it follows that between the two pauses there must be a flight indexed by $t_x$ of length $j-i+n$.
$$t_x =  t_{i+n} < ... < t_{j} |t_x \in F$$
We define pause locations $G(t_a), G(t_b) | t_a, t_b \in P$ as locations where an individual spends an extended amount of time in the same space (e.g. school, home, work, train station, barber shop, bar, gym). Importantly, our model assumes period and cyclic human movement such that there are many pauses $t_{a1},t_{a2},...,t_{an}$ such that $G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})$. Moreover,it is possible for two consecutive pauses $t_a$ and $t_b$ to share a location $G(t_a) = G(t_b)$ such that $t_a \ne t_b$. For example, if the individual leaves home for a run and returns home without stopping anywhere else.

@@

## Personalised Map Matched Imputation algorithm

@@

1. *Map building*: Extract from measurements $D$ all pause locations and path locations to create a personalised map consisting of discrete bins.
2. *Binning*: Assign each measurement $D$ to a unique discrete location. 
3. *Imputing*: Use a classification method to predict missing measurements based on all the available information.

@@

Following @wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete locations, which we will call location bins. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007]. We will focus on methods which explicitly deal with mobility patterns extracted from smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

@@

1. The next measurement $D_{t+1}$ is within $t_{\text{Pause lim}}$ amount of seconds (so it is not missing)
2. The next measurement $D_{t+1}$ is within $d_{\text{Pause lim}}$ meters.
3. The duration of the pause is more than $\delta_{\text{Pause lim}}$ seconds.
4. Let the measurements of a possible pause which fit the aforementioned criteria be $X = D_{t,t+1,...,t+n}$. Let $\bar{x}$ be the mean coordinates of $X$, let $x_{\text{dist}} \in X$ be the furthest away points from $\bar{x}$ and $\bar{a}$ be the the mean accuracy of $X$. Then $X$ is a pause if and only if the distances between $\bar{x}$ and $x_{\text{dist}}$ are $\leq 2 \times \bar{a}$.

The resulting set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a bin.

@@

1. Take all measurements which are not pauses, filter them based on an accuracy threshold $a_{\text{Path Lim}}$.
2. Create a distance matrix for all remaining measurements $D_t \in F$ and hierarchically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single flight bin.

At this point all empirically observed path bins and pause bins are extracted. However, there may be some overlap between pause bins and path bins. Thus, the bins are clustered again, such that the pause bins retain priority. This means that if a pause bin and a path bin are within less than $d$ meters of each other, the path bin is removed. The reasoning for this is that the threshold for not being in a pause bin should be higher, as individuals spend the majority of time at a pause bin. The end result is a discretised map which contains pause and flight bins based on the entire log history of the individual.

@@

PMMI uses a similar logic, but without using any external geographic database. The key modification in PMMI is that whilst @wu_spatial-temporal-semantic_2017 uses an external map, we use the total location history of the individual to create a personalised map. This map is subsequently used to bin measurements. This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long. To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_{ab}$.

Thus, all measurements $D$ are then assigned to a discrete bin on the personal map. This includes previous measurements which were discarded from the map building exercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$. In this implementation we used a simple assigning function, whereby the measurements where assigned to the nearest bin.
